Submodel descriptions: Radiative transfer

### 2019-07-16 analysis/paper/submodel-rtm.Rmd
Add lots of info to RTM submodel description

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
stopifnot(
  requireNamespace("purrr", quietly = TRUE),
  requireNamespace("gsl", quietly = TRUE)
)
```

@@

Following @clm45_note, forward- ($\nu$) and backscattering of leaves or stems ($\nu$) are defined as a function of reflectance ($R$) and transmittance ($T$) of the corresponding materials:

@@

```{r g-function, echo = FALSE, fig.cap = gfun_cap}
gfun_cap <- "Leaf projection as a function of leaf orientation factor."
phi1_f <- function(orient) {
  0.5 - 0.633 * orient - 0.33 * orient ^ 2
}
phi2_f <- function(orient, phi1 = phi1_f(orient)) {
  0.877 * (1 - 2 * phi1)
}
gfunction <- function(theta, orient, theta_rad = theta * pi / 180) {
  phi1 <- phi1_f(orient)
  phi2 <- phi2_f(phi1 = phi1)
  mu <- cos(theta_rad)
  phi1 + phi2 * mu
}
max_theta <- 89
orient <- c(-0.5, -0.25, 0, 0.25, 0.5)
colors <- colorRampPalette(c("red", "black", "blue"))(length(orient))
plot(c(-90, 90), c(0, 0.9), type = "n",
     xlab = "Zenith angle (degrees)",
     ylab = expression(G(psi)))
legend("bottom", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
invisible(purrr::map2(
    orient, colors,
    ~curve(gfunction(x, .x), -max_theta, max_theta,
           col = .y, add = TRUE)
))
```

```{r kfunction, echo = FALSE, fig.cap = kcap}
kcap <- "K function"
kfunction <- function(theta, orient, theta_rad = theta * pi / 180) {
  gfunction(theta, orient, theta_rad) / cos(theta_rad)
}
plot(c(-70, 70), c(0.2, 1.8), type = "n",
     xlab = "Zenith angle (degrees)",
     ylab = expression(K(psi)))
legend("top", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
invisible(purrr::map2(
  orient, colors,
  ~curve(kfunction(x, .x), -70, 70,
         col = .y, add = TRUE)
))
```

@@

```{r}
mu_bar <- function(orient) {
  phi1 <- phi1_f(orient)
  phi2 <- phi2_f(phi1 = phi1)
  x <- 1 - phi1 / phi2 * log(1 + phi2 / phi1)
  out <- x / phi2
  out[orient == 0] <- 1
  out
}
curve(mu_bar(x), -0.5, 0.5,
      xlab = "Orient factor",
      ylab = expression(bar(mu)))
```

@@

$$
a_s(\psi) = \frac{1}{2}
\frac{G(\psi)}{\cos\psi \phi2 + G(\psi)}
\left(
1 - 
\frac{\cos\psi \phi1}{\cos\psi + G(\psi)}
\ln\left(
\frac{\cos\psi \phi1 + \cos\psi \phi2 + G(\psi)}{\cos\psi \phi1}
\right)
\right)
$$

@@

@@

@@

@@

The _direct_ ("beam") transmittance fraction of canopy layer $i$ at solar zenith angle direction $\psi$ ($\tau(\psi)_i$) is defined as:

$$ \tau(\psi)_i = \exp\left( -K(\psi)_i L_i \right) $$

```{r int-direct, fig.cap = idcap}

idcap <- "Direct transmittance fraction, as a function of LAI"
tau_direct <- function(psi, L, orient) {
  exp(-kfunction(psi, orient) * L)
}
Lmax <- 5
plot(c(0, Lmax), c(0, 1), type = "n",
     xlab = "Leaf area index",
     ylab = "Direct tranmittance fraction")
legend("topright", c("-0.5", "0", "0.5"),
       col = c("red", "black", "blue"), lty = 1,
       title = "Orient factor")
legend("top", c("0", "30", "60"), lty = 1:3,
       title = "Zenith angle")
curve(tau_direct(0, x, 0), 0, Lmax, add = TRUE)
curve(tau_direct(0, x, -0.5), 0, Lmax, add = TRUE, col = "red")
curve(tau_direct(0, x, 0.5), 0, Lmax, add = TRUE, col = "blue")
curve(tau_direct(30, x, 0), 0, Lmax, add = TRUE, lty = 2)
curve(tau_direct(30, x, -0.5), 0, Lmax, add = TRUE, lty = 2, col = "red")
curve(tau_direct(30, x, 0.5), 0, Lmax, add = TRUE, lty = 2, col = "blue")
curve(tau_direct(60, x, 0), 0, Lmax, add = TRUE, lty = 3)
curve(tau_direct(60, x, -0.5), 0, Lmax, add = TRUE, lty = 3, col = "red")
curve(tau_direct(60, x, 0.5), 0, Lmax, add = TRUE, lty = 3, col = "blue")

where $L_i$ is the layer's leaf area index (technically, total area index), and $K(\psi)_i$ is the directional extinction coefficient (see "Directional scattering and leaf angle distribution").

@@

tau_diffuse <- function(L, orient) {
  phi1 <- phi1_f(orient)
  phi2 <- phi2_f(phi1 = phi1)
  ext_diff1 <- phi1 * L
  ext_diff2 <- phi2 * L
  -exp(-ext_diff1 - ext_diff2) *
    (ext_diff1^2 * exp(ext_diff1) * gsl::expint_Ei(-ext_diff1) +
       ext_diff1 - 1)
}
plot(c(0, Lmax), c(0, 1), type = "n",
     xlab = "Leaf area index",
     ylab = "Diffuse transmittance fraction")
legend("topright", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
purrr::walk2(
  orient, colors,
  ~curve(tau_diffuse(x, .x), 0, Lmax, add = TRUE, col = .y)
)

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
stopifnot(
  requireNamespace("purrr", quietly = TRUE),
  requireNamespace("gsl", quietly = TRUE)
)
library(fortebaseline)
```

@@

Following @clm45_note, forward- ($\nu$) and backscattering of leaves or stems ($\omega$) are defined as a function of reflectance ($R$) and transmittance ($T$) of the corresponding materials:

@@

```{r g-function, echo = FALSE, fig.cap = gfun_cap}
gfun_cap <- "Leaf projection as a function of leaf orientation factor."
max_theta <- 89
orient <- c(-0.5, -0.25, 0, 0.25, 0.5)
colors <- colorRampPalette(c("red", "black", "blue"))(length(orient))
plot(c(-90, 90), c(0, 0.9), type = "n",
     xlab = "Zenith angle (degrees)",
     ylab = expression(G(psi)))
legend("bottom", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
invisible(purrr::map2(
    orient, colors,
    ~curve(gfunction(x, .x), -max_theta, max_theta,
           col = .y, add = TRUE)
))
```

```{r kfunction, echo = FALSE, fig.cap = kcap}
kcap <- "K function"
plot(c(-70, 70), c(0.2, 1.8), type = "n",
     xlab = "Zenith angle (degrees)",
     ylab = expression(K(psi)))
legend("top", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
invisible(purrr::map2(
  orient, colors,
  ~curve(kfunction(x, .x), -70, 70,
         col = .y, add = TRUE)
))
```

@@

```{r}
curve(mu_bar(x), -0.5, 0.5,
      xlab = "Orient factor",
      ylab = expression(bar(mu)))
```

@@

$$
a_s(\psi) = \frac{1}{2}
\frac{G(\psi)}{\cos\psi \phi_2 + G(\psi)}
\left(
1 - 
\frac{\cos\psi \phi_1}{\cos\psi + G(\psi)}
\ln\left(
\frac{\cos\psi \phi_1 + \cos\psi \phi_2 + G(\psi)}{\cos\psi \phi_1}
\right)
\right)
$$

@@

```{r single-scattering-albedo, fig.cap = ssa_cap}
ssa_cap <- paste("Single-scattering albedo as a function of ",
                 "radiation angle and leaf orientation factor.")
plot(c(-90, 90), c(0, 1.3), type = "n",
     xlab = expression("Zenith angle" ~ (psi * degree)),
     ylab = expression("Single-scattering albedo" ~ (a[s])))
curve(single_scatter_albedo(x, 0), -89, 89, add = TRUE)
curve(single_scatter_albedo(x, 0.25), -89, 89, col = "blue", add = TRUE)

curve(single_scatter_albedo(x, -0.25), -89, 89, col = "red", add = TRUE)
legend("topright", c("-0.25", "0", "0.25"), title = "Orient factor",
       col = c("red", "black", "blue"), lty = 1)
```

```{r beta0, fig.cap = beta0_cap}
beta0_cap <- paste(
  "Backscatter for direct radiation",
  "as a function of radiation angle",
  "and leaf orientation factor."
)
plot(c(-90, 90), c(0.4, 4), type = "n",
     xlab = expression("Zenith angle" ~ (psi * degree)),
     ylab = expression("Direct radiation backscatter" ~ (beta[0])))
curve(beta0(x, 0), -89, 89, add = TRUE)
curve(beta0(x, 0.25), -89, 89, col = "blue", add = TRUE)
curve(beta0(x, -0.25), -89, 89, col = "red", add = TRUE)
legend("topright", c("-0.25", "0", "0.25"), title = "Orient factor",
       col = c("red", "black", "blue"), lty = 1)
```

@@

```{r int-direct, fig.cap = idcap}
idcap <- "Direct transmittance fraction, as a function of LAI"
Lmax <- 5
plot(c(0, Lmax), c(0, 1), type = "n",
     xlab = "Leaf area index",
     ylab = "Direct tranmittance fraction")
legend("topright", c("-0.5", "0", "0.5"),
       col = c("red", "black", "blue"), lty = 1,
       title = "Orient factor")
legend("top", c("0", "45"), lty = 1:2,
       title = "Zenith angle")
curve(tau_direct(x, 0, 0), 0, Lmax, add = TRUE)
curve(tau_direct(x, 0, -0.5), 0, Lmax, add = TRUE, col = "red")
curve(tau_direct(x, 0, 0.5), 0, Lmax, add = TRUE, col = "blue")
curve(tau_direct(x, 45, 0), 0, Lmax, add = TRUE, lty = 2)
curve(tau_direct(x, 45, -0.5), 0, Lmax, add = TRUE, lty = 2, col = "red")
curve(tau_direct(x, 45, 0.5), 0, Lmax, add = TRUE, lty = 3, col = "blue")
```

@@

The _direct_ ("beam") transmittance fraction of canopy layer $i$ at solar zenith angle direction $\psi$ ($\tau(\psi)_i$) is defined above.

@@

@@

plot(c(0, Lmax), c(0, 1), type = "n",
     xlab = "Leaf area index",
     ylab = "Diffuse transmittance fraction")
legend("topright", as.character(orient),
       title = "Orient factor",
       lty = "solid",
       col = colors)
purrr::walk2(
  orient, colors,
  ~curve(ms_tau_diffuse(x, .x), 0, Lmax, add = TRUE, col = .y)
)
